Structural analysis

### 2025-03-19 workflow/notebooks/draft.Rmd
added new figures and revision CXX analysis

models <- names(palettes_model)

@@

```{r read_homo}
fs <- read_delim(glue(homodir, "/", params_dataset$seed, "_fs.tsv"), 
                 col_names = fs_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue)) %>%
  left_join(taxidmap, by = "target")

@@

```{r merge_homo}
df <- full_join(blast, fs,
                by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
    filter(query!=target) %>% 
  left_join(self, by="query") %>%
  mutate(pident_Foldseek = 100*pident_Foldseek, 
         qcov_Foldseek = 100*qcov_Foldseek,
         singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
                               is.na(evalue_Blast) ~ "only Foldseek", 
                               .default = "Common"))

@@

```{r cath}
CATH_df <- read_delim(paste0(params$data_dir, "cath/", tax$Proteome_ID, "_cath.tsv"),
                      delim = "\t", id = "proteome", show_col_types = FALSE) %>%
  mutate(proteome = gsub("_cath.tsv", "", basename(proteome))) %>%
         # n_3d_domains = str_count(Gene3D, ";"),
         # n_Pfam_domains = str_count(Pfam, ";")
  mutate(target=paste0("AF-", Entry, "-F1")) %>% 
  rename("CATH"="Gene3D") %>% 
  select(target, Pfam, CATH, -proteome, -Entry) %>% 
  mutate(CAT = gsub("(\\d+\\.\\d+\\.\\d+)\\.\\d+", "\\1", CATH),
         CA = gsub("(\\d+\\.\\d+)\\.\\d+", "\\1", CAT),
         C = gsub("\\.\\d+", "\\1", CA))

@@

```{r fig1}
fig1a <- df %>%
  group_by(singleton) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(prop=n/sum(n), Filtering="Before") %>%
  bind_rows(df_filt %>%
            group_by(singleton) %>%
            summarise(n = n()) %>%
            ungroup() %>%
            mutate(prop=n/sum(n), Filtering="After")) %>%
  mutate(Filtering=factor(Filtering, labels = c("After", "Before"), ordered = T),
         singleton=case_when(singleton=="only Blast"~"Only Bp", 
                             singleton=="only Foldseek" ~ "Only Fs",
                             .default = singleton)) %>%
  ggplot(aes(singleton, n, fill=singleton)) +
  geom_bar(aes(alpha=Filtering), stat = "identity", position = position_dodge2(reverse=TRUE)) +
  scale_fill_manual(values = palette_singleton_s, guide="none") +
  scale_alpha_manual(values = c(1,.6)) +
  scale_y_continuous(expand = c(0,0), labels = scales::label_number(scale = 1e-6)) +
  labs(fill="Class", x="Class", y="Milions of aligned protein pairs") +
  theme(legend.position = "bottom", legend.box="vertical")

@@

```{r fig2}
eval_p <- ggplot(df, aes(-log10(evalue_Blast), -log10(evalue_Foldseek))) + 
  stat_bin_2d(breaks=seq(0,20,0.4)) + 
  scale_fill_gradientn(colors=wespal) +
  ggpubr::stat_cor(label.x.npc = .025) +
  coord_cartesian(expand = 0) +
  labs(x="-log10(Evalue BlastP)",y="-log10(Evalue Foldseek)") +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(legend.position = "none", panel.grid = element_blank()) 
eval_p <- add_marginal(mutate(df, 
                              evalue_Blast=-log10(evalue_Blast), 
                              evalue_Foldseek=-log10(evalue_Foldseek)), 
                       eval_p, "evalue_Blast", "evalue_Foldseek", "singleton", palette_singleton_p)

@@

# Loop through each column name and apply a function
# jaccard_df_mono <- map_dfr(dbs, ~ {
#   col_name <- .x
#   select(df_filt, query, target, singleton) %>% 
#     left_join(select(ortho_dbs, target, col_name), by=c("query"="target")) %>% 
#     left_join(select(ortho_dbs, target, col_name), by=c("target")) %>% 
#     rename("query_set"=paste0(col_name, ".x"), "target_set"=paste0(col_name, ".y")) %>% 
#     mutate(n_domain_q = str_count(pattern = "\\;", query_set)) %>% 
#     # Get only monodomain queries
#     filter(n_domain_q==1) %>% 
#     filter(!is.na(query_set), !is.na(target_set)) %>% 
#     mutate(
#       jaccard = map2_dbl(strsplit(query_set, ";"), strsplit(target_set, ";"),
#                          ~ length(intersect(.x, .y)) / length(union(.x, .y)))
#     ) %>% 
#     group_by(query, singleton) %>% 
#     summarise(avg_Jaccard = mean(jaccard))  %>% 
#     mutate(db=col_name)
# })
# 
# jaccard_df_NA <- map_dfr(dbs, ~ {
#   col_name <- .x
#   select(df_filt, query, target, singleton) %>%
#     left_join(select(ortho_dbs, target, col_name), by=c("query"="target")) %>%
#     left_join(select(ortho_dbs, target, col_name), by=c("target")) %>%
#     rename("query_set"=paste0(col_name, ".x"), "target_set"=paste0(col_name, ".y")) %>%
#     # If target has na no we count mismatch
#     filter(!is.na(query_set)) %>%
#     mutate(
#       jaccard = map2_dbl(strsplit(query_set, ";"), strsplit(target_set, ";"),
#                          ~ length(intersect(.x, .y)) / length(union(.x, .y)))
#     )  %>%
#     group_by(query, singleton) %>%
#     summarise(avg_Jaccard = mean(jaccard)) %>%
#     mutate(db=col_name)
# 
# })

@@

# p_dbs_mono <-  ggplot(jaccard_df_mono, aes(avg_Jaccard, 1, color=singleton, 
#                                 linetype=db, alpha=db)) +
#   stat_ecdf(pad=FALSE) +
#   scale_color_manual(values=palette_singleton_p) +
#   scale_linetype_manual(values = c(2,2,2,2,1)) +
#   scale_alpha_manual(values = c(0.4,0.6,0.8,1,1)) +
#   coord_fixed() +
#   scale_y_continuous(expand = expansion(0.02,0)) +  
#   scale_x_continuous(expand = expansion(0.02,0)) +  
#   theme_bw(base_family = "Aptos") +
#   labs(title="Only Monodomain queries", x="Mean Jaccard Index", y="ECDF", 
#        alpha="DB", linetype="DB", color="Target sets")
# 
# p_dbs_NA <-  ggplot(jaccard_df_NA, aes(avg_Jaccard, 1, color=singleton,
#                                 linetype=db, alpha=db)) +
#   stat_ecdf(pad=FALSE) +
#   scale_color_manual(values=palette_singleton_p) +
#   scale_linetype_manual(values = c(2,2,2,2,1)) +
#   scale_alpha_manual(values = c(0.4,0.6,0.8,1,1)) +
#   coord_fixed() +
#   scale_y_continuous(expand = expansion(0.02,0)) +
#   scale_x_continuous(expand = expansion(0.02,0)) +
#   theme_bw(base_family = "Aptos") +
#     labs(title="NA=mismatch", x="Mean Jaccard Index", y="ECDF", 
#        alpha="DB", linetype="DB", color="Target sets")

# jaccard_plot <- (p_dbs_NA | p_dbs_mono) + 
#   plot_layout(guides = "collect", 
#               axis_titles = "collect") & theme(legend.position = 'bottom')

@@

ggsave("draft/figures/Fig2.pdf", fig2, width = 8, height = 7, device = cairo_pdf)
# ggsave("draft/figures/Fig_Supp_CATH.pdf", jaccard_plot, width = 8, height = 4, device = cairo_pdf)
```

@@

```{r read_trees}
models_draft <- setdiff(models, c("QT", "FTPY", "AF", "LLM"))
df_trees <- read_delim(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/trees/", 
                              params_dataset$seed, "_unrooted_trees.txt"),
                       show_col_types = FALSE,
                       col_names = c("gene", "target", "alphabet", "model", "tree")) %>% 
  mutate(model=factor(model, levels=models)) %>% 
  filter(!is.na(tree), model %in% models_draft) %>% 
  # some FM trees have negative branch lengths
  mutate(tree = gsub(":-", ":", tree)) 
# 722*8*4=23104 trees

@@

# Reconciliation ranger+TCS scores
# glue("results/phylogeny/", params_dataset$phylo_dataset, "/reco/", params_dataset$seed, "_notung.tsv")
reco <- read_delim(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/reco/", 
                              params_dataset$seed, "_notung.tsv"), 
                   show_col_types = FALSE, 
                   col_names = c("gene", "target", "alphabet", "model", "dups", "losses")) %>% 
  left_join(n_tips) %>% 
  left_join(var_df) %>% 
  mutate(model=factor(model, levels=models),
         n_events = (dups+losses)/n_tips) %>% 
  filter(model %in% models_draft) %>% 
  rename("targets"="target")
```

@@

apro_files <- list.files("test/flashforward/reco/", 
                         pattern = "^apro*", full.names = TRUE)

@@

@@

```{r}
# find results/phylogeny/hsap_1kseeds/seeds/UP000005640/ -name "*.ids" -print0 | while IFS= read -r -d '' file; do  awk -v fname="$(basename "$file" | sed 's/.ids//' | sed 's/_/\t/')"  '{print $0 "\t" fname}' "$file"; done > test/paranoia/ids_by_set.txt
df_ids <- read_delim("draft/figures/ids_by_set.txt", 
                 col_names = c("target", "query", "set")) %>% 
  filter(query %in% aln_seeds)

@@

```{r}
lm <- lm(dups+losses ~ n_tips, data = reco)
reco$res <- lm$residuals

@@

>>>>

@@

```{r read_homo, eval=FALSE}
fs <- read_delim(glue(homodir, "/", params_dataset$seed, "_fs.tsv"), 
                 col_names = fs_columns, show_col_types = FALSE) %>% 
  # filter(query %in% seeds) %>%
  mutate(evalue = ifelse(evalue==0, 1e-180, evalue)) %>%
  left_join(taxidmap, by = "target")

@@

```{r merge_homo, eval=FALSE}
df <- full_join(blast, fs,
                by = c("query", "target"), suffix=c("_Blast", "_Foldseek")) %>%
    filter(query!=target) %>% 
  left_join(self, by="query") %>%
  mutate(pident_Foldseek = 100*pident_Foldseek, 
         qcov_Foldseek = 100*qcov_Foldseek,
         singleton = case_when(is.na(evalue_Foldseek) ~ "only Blast",
                               is.na(evalue_Blast) ~ "only Foldseek", 
                               .default = "Common"))

@@

```{r cath, eval=FALSE}
CATH_df <- read_delim(paste0(params$data_dir, "cath/", tax$Proteome_ID, "_cath.tsv"),
                      delim = "\t", id = "proteome", show_col_types = FALSE) %>%
  mutate(proteome = gsub("_cath.tsv", "", basename(proteome))) %>%
         # n_3d_domains = str_count(Gene3D, ";"),
         # n_Pfam_domains = str_count(Pfam, ";")
  mutate(target=paste0("AF-", Entry, "-F1")) %>% 
  rename("CATH"="Gene3D") %>% 
  select(target, Pfam, CATH, -proteome, -Entry) %>% 
  mutate(CAT = gsub("(\\d+\\.\\d+\\.\\d+)\\.\\d+", "\\1", CATH),
         CA = gsub("(\\d+\\.\\d+)\\.\\d+", "\\1", CAT),
         C = gsub("\\.\\d+", "\\1", CA))

@@

```{r fig1, eval=FALSE}
fig1a <- df %>%
  group_by(singleton) %>%
  summarise(n = n()) %>%
  ungroup() %>%
  mutate(prop=n/sum(n), Filtering="Before") %>%
  bind_rows(df_filt %>%
            group_by(singleton) %>%
            summarise(n = n()) %>%
            ungroup() %>%
            mutate(prop=n/sum(n), Filtering="After")) %>%
  mutate(Filtering=factor(Filtering, labels = c("After", "Before"), ordered = T),
         singleton=case_when(singleton=="only Blast"~"Only Bp", 
                             singleton=="only Foldseek" ~ "Only Fs",
                             .default = singleton)) %>%
  ggplot(aes(singleton, n, fill=singleton)) +
  geom_bar(aes(alpha=Filtering), stat = "identity", position = position_dodge2(reverse=TRUE)) +
  scale_fill_manual(values = palette_singleton_s, guide="none") +
  scale_alpha_manual(values = c(1,.6)) +
  scale_y_continuous(expand = c(0,0), labels = scales::label_number(scale = 1e-6)) +
  labs(fill="Class", x="Class", y="Milions of aligned protein pairs") +
  theme(legend.position = "bottom", legend.box="vertical")

@@

```{r fig2, eval=FALSE}
eval_p <- ggplot(df, aes(-log10(evalue_Blast), -log10(evalue_Foldseek))) + 
  stat_bin_2d(breaks=seq(0,20,0.4)) + 
  scale_fill_gradientn(colors=wespal) +
  ggpubr::stat_cor(label.x.npc = .025) +
  coord_cartesian(expand = 0) +
  labs(x="-log10(Evalue BlastP)",y="-log10(Evalue Foldseek)") +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_continuous(limits = c(0, 20)) +
  theme(legend.position = "none", panel.grid = element_blank()) 
eval_p <- add_marginal(mutate(df, 
                              evalue_Blast=-log10(evalue_Blast), 
                              evalue_Foldseek=-log10(evalue_Foldseek)), 
                       eval_p, "evalue_Blast", "evalue_Foldseek", "singleton", palette_singleton_p)

@@

@@

@@

ggsave("draft/figures/Fig2.pdf", fig2, width = 8, height = 7, device = cairo_pdf)
```

@@

```{r read_trees}
models_draft <- setdiff(models, c("QT", "FTPY", "AF", "LLM"))

df_trees <- read_delim(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/trees/", 
                              params_dataset$seed, "_unrooted_trees.txt"),
                       show_col_types = FALSE,
                       col_names = c("gene", "target", "alphabet", "model", "tree")) %>% 
  mutate(model=factor(model, levels=models)) %>% 
  filter(!is.na(tree)) %>% #, model %in% models_draft) 
  # some FM trees have negative branch lengths
  mutate(tree = gsub(":-", ":", tree)) 
# 722*8*4=23104 trees

@@

# Reconciliation ranger+TCS scores
# glue("results/phylogeny/", params_dataset$phylo_dataset, "/reco/", params_dataset$seed, "_notung.tsv")
reco <- read_delim(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/reco/", 
                              params_dataset$seed, "_notung.tsv"), 
                   show_col_types = FALSE, 
                   col_names = c("gene", "target", "alphabet", "model", "dups", "losses")) %>% 
  left_join(n_tips) %>% 
  left_join(var_df) %>% 
  mutate(model=factor(model, levels=models),
         n_events = (dups+losses)/n_tips) %>% 
  # filter(model %in% models_draft) %>% 
  rename("targets"="target")

apro_files <- list.files(paste0(params$outdir, 'phylogeny/',
                              params_dataset$phylo_dataset,  "/reco/apro"), 
                        pattern = "*support.nwk", full.names = TRUE)

save.image(file = "workflow/notebooks/eda_cache/until_dftree.Rdata")

knitr::knit_exit()
```

@@

@@

```{r fig3_sig}
dl_sig <- reco %>% 
  rename("y"="n_events") %>% 
  plot_stat() + 
  labs(y="DL")

disco_sig <- disco_rf %>% 
  mutate(model=factor(model, levels=models)) %>% 
  filter(nsps>length(sptree$tip.label)/2, model %in% models_draft) %>%
  # separate(bn, c("seed", "targets", "alphabet", "model")) %>% 
  left_join(df_model) %>% 
  rename("y"="rf", "targets"="target") %>% 
  plot_stat() + 
  labs(y="RF")

apro_sig <- trees_df %>% 
  select(-y) %>% 
  left_join(df_model) %>% 
  rename("y"="q1") %>% 
  plot_stat() + 
  labs(y="q1")

(disco_sig | apro_sig | dl_sig) + 
  patchwork::plot_layout(guides = "collect") & 
  theme(legend.position = "bottom")
```

@@

```{r paralogy_phylome}
# find results/phylogeny/hsap_1kseeds/seeds/UP000005640/ -name "*.ids" -print0 | while IFS= read -r -d '' file; do  awk -v fname="$(basename "$file" | sed 's/.ids//' | sed 's/_/\t/')"  '{print $0 "\t" fname}' "$file"; done > test/paranoia/ids_by_set.txt
df_ids <- read_delim("draft/figures/ids_by_set.txt", 
                 col_names = c("target", "query", "set")) %>% 
  filter(query %in% aln_seeds)

@@

```{r heteroscedasticity}
lm <- lm(dups+losses ~ n_tips, data = reco)
reco$res <- lm$residuals

@@

### CXX models

```{r cxx_models}
times_df <- read_delim("tmp.txt", 
                       col_names = c("basename","s","h:m:s","max_rss",
                                     "max_vms","max_uss","max_pss",
                                     "io_in","io_out","mean_load","cpu_time")) %>% 
  mutate(max_rss_gb = max_rss/1024, basename=gsub(".txt", "", basename)) %>% 
  separate(basename, c("seed", "gene", "targets", "alphabet", "model"), sep = "_") %>% 
  select(gene, targets, model, s, `h:m:s`, max_rss_gb)

plot_time_cxx <- times_df %>% 
  # filter() %>% 
  ggplot(aes(fct_reorder(model, s), s/3600)) + 
  geom_boxplot(outlier.size = .1) +
  # scale_y_log10() + 
  scale_y_continuous(transform = "log10")

### Query coverage and BlastP singletons

```{r qcov_blastpsingletons}
pfam_jaccard <- df %>% 
  # head(100000) %>% 
  select(query, target, qcov_Blast, qcov_Foldseek, singleton) %>% 
  left_join(select(ortho_dbs, target, Pfam), by=c("query"="target")) %>% 
  left_join(select(ortho_dbs, target, Pfam), by=c("target")) %>% 
  rename("query_set"=paste0("Pfam", ".x"), "target_set"=paste0("Pfam", ".y")) %>% 
  # If any of query or target has na no we dont count
  filter(!is.na(query_set), !is.na(target_set)) %>% 
  mutate(
    jaccard = map2_dbl(strsplit(query_set, ";"), strsplit(target_set, ";"),
                       ~ length(intersect(.x, .y)) / length(union(.x, .y)))
  )

pfam_jaccard %>% 
  mutate(qcov = rowMeans(select(., qcov_Blast, qcov_Foldseek), na.rm = TRUE)) %>% 
  mutate(disc=cut(qcov, breaks = seq(0,100,25))) %>% 
  ggplot(aes(jaccard, disc, color=singleton)) + 
  geom_boxplot() + 
  scale_color_manual(values=palette_singleton_p)

query_pfam <- pfam_jaccard %>% 
  # filter(query=="AF-P06681-F1") %>% 
  group_by(query) %>% 
  summarise(#all_domains = paste0(target_set, collapse = ""),
            q_domains = paste0(unique(query_set), collapse = "")) %>% 
  separate(q_domains, letters, sep = ";")  %>% 
  pivot_longer(cols = !c(query), values_drop_na = T)  %>% 
  filter(value!="") %>% 
  select(-name)

top_pfam <- pfam_jaccard %>% 
  # filter(query=="AF-P06681-F1") %>%
  select(query, target_set)  %>% 
  left_join(query_pfam, relationship = "many-to-many") %>% 
  mutate(is_in = str_detect(target_set, value)) %>%
  group_by(query, value) %>% 
  summarise(n = sum(is_in)) %>% 
  slice_max(n, with_ties = FALSE)

prop_topdomain <- pfam_jaccard %>% 
  # filter(query=="AF-A0A024RCN7-F1") %>% 
  select(query, target_set, singleton) %>% 
  left_join(top_pfam) %>% 
  mutate(is_in = str_detect(target_set, value)) %>% 
  group_by(query, singleton) %>% 
  mutate(n_group = n()) %>% 
  summarise(n = sum(is_in)/mean(n_group))

prop_topdomain %>% 
  ggplot(aes(n, singleton)) + 
  geom_boxplot() + 
  labs(x="% with top domain")
```
